---
title: "Story of analysis - IMF dashboard analysis"
author: Max Franke
date: April 28, 2020
output: pdf_document
header-includes:
   - \usepackage{animate}
---
Course: CIS-546 DATA VISUALIZATION

Term: Spring/T2

Data: The IMF publishes a range of time series data on IMF lending, exchange rates and other economic and financial indicators.

# Approach of the analysis

This analysis aims to visualize the collected data to create a dashobard which shows the data in a structured and visualized way.

First, the data is collected by calling the API for IMF (package: 'imfr': 

https://cran.r-project.org/web/packages/imfr/imfr.pdf)

# Data presenation

In this section the data in its raw version is presented. 

```{r echo = FALSE}
## Clean Workspace
rm(list = ls())
```

```{r echo = FALSE}
## Install packages
#install.packages("tidyverse")
#install.packages("knitr")
#install.packages("rmarkdown")
#install.packages("tinytex")
#install.packages("dplyr")
#install.packages("kableExtra")
```

```{r warning=FALSE, message=FALSE, echo = FALSE}
## Load libraries
library(tidyverse)
library(knitr)
library(rmarkdown)
#tinytex::install_tinytex()
library(tinytex)
library(dplyr)
library(kableExtra)
options(knitr.table.format = "latex")
devtools::install_github('yihui/knitr')
library(knitr)
library(shiny)
library(shinydashboard)
library(tidyverse)
library(dplyr)
library(readxl)
library(reshape)
library(gganimate)
library(gifski)
library(png)
library(plotly)
```

```{r warning=FALSE, message=FALSE, echo = FALSE}
## Load data of stocks with headlines and description
setwd("/Users/MaxFranke/Desktop/05_Big Data Analytics/04_Classes/04 SP:Term2/CIS-546 DATA VISUALIZATION/05_TermProject/TermProject/01_Data")
gdp <- read.csv("WEOOct2019all.csv")
weo <- read.csv("WEO_Data.csv")
```

```{r warning=FALSE, message=FALSE, echo = FALSE}
## First look at the data
knitr::kable(head(gdp[,c(4,5,40:44)]), "latex", 
             booktabs = T, 
             caption = "Extract from the full data set (not all variables)") %>%
  kable_styling(full_width = TRUE) %>%
  column_spec(column = 2, width = "3cm")
  
```


## Description of data

The datasets show years as different columns and the dimensions are wrong, which means that data wrangling and processing has to be done:

1. Change the column names from Year by deleting the first character "X"

2. Select the important columns (e.g. delte the WEO.Country.Code)

3. Use "melt" in order to summarise all years columns in one column and use "WEO.Subject.Code", and "Country" as an ID

4. Change the summarized values from factor to numeric values

5. Change the Year column to format "Year"

```{r warning=FALSE, message=FALSE, echo = FALSE}
names(gdp)[10:54] <- substring(names(gdp)[10:54], 2)
GDP_development <- subset(gdp, WEO.Subject.Code == "NGDP_R")[,c(3,4,10:50)]

GDP_development <- melt(subset(gdp, WEO.Subject.Code == "NGDP_R")[,c(3,4,10:50)], id = c("WEO.Subject.Code", "Country"))

GDP_development$value <- as.numeric(GDP_development$value)
#GDP_development[!which(GDP_development$variable %in% c(2024, 2023, 2022, 20221)),]
GDP_development$variable <- as.Date(paste0(as.character(GDP_development$variable), '-01-01'))
names(GDP_development) <- c("WEO.Subject.Code", "Country", "Year", "GDP")

knitr::kable(head(GDP_development), "latex", 
             booktabs = T, 
             caption = "Head of transformed data with GDP") %>%
  kable_styling(full_width = TRUE)
```


# Idea of the dashboard

The dashboard is structured into three areas (channels):

1. National Account

2. Trade

3. People

In the first area, following variables are included:

 - Gross domestic product, constant prices 
 - Gross domestic product, constant prices (percent changes)
 - Inflation, average consumer prices
 - Total investment 

In the second area, following variables are included:

 - Volume of imports of goods and services
 - Volume of exports of goods and services
 - Volume of Imports of goods
 - Volume of exports of goods
 
In the third area, following variables are included:

 - Unemployment rate
 - Employment
 - Population
 - Current account balance

In the following chapters, the visualization will be produced, which will be impelemented into the the dashboard with shiny.

The dashboard can be found with the following link:

https://max-franke.shinyapps.io/IMF_analysis/

# National Account

```{r, fig.show='animate'}
library(gganimate)
names(weo)[4:44] <- substring(names(weo)[4:44], 2)

for (i in 4:19) {
weo[,i] <- as.numeric(levels(weo[,i])[weo[,i]])  
}

weo_table <- melt(subset(weo, Subject.Descriptor == "Gross domestic product, constant prices" & Units == "Percent change")[,c(1,2,4:44)], id = c("Country.Group.Name", "Subject.Descriptor"))
weo_table$variable <- round(as.numeric(levels(weo_table$variable)[weo_table$variable]),0)
weo_table$hi_lo <- if_else(weo_table$value > 0, "Above", "Below")

ggplot(data = subset(weo_table, Country.Group.Name %in% c("World", "European Union", "Middle East and Central Asia")), 
           mapping = aes(x = variable, y = value, fill = hi_lo)) +
  geom_col(position = "dodge2") +
  facet_wrap(~Country.Group.Name, nrow = 1, ncol = 3) +
  guides(fill = FALSE) +
  scale_fill_manual(values = c( "#009E73", "#D55E00")) +
  labs(title = "Development of GDP for the World",
       x = NULL,
       y = "Percent change",
       subtitle = "Date: {closest_state} (Min: 1980, Max: 2020)") +
  transition_states(variable, wrap = FALSE) +
  shadow_mark() +
  enter_grow() +
  enter_fade()
```





























































